---
title: "Untitled"
author: "John Stansfield"
date: "September 13, 2017"
output: html_document
---

# test p-value adjustment

```{r}
# power-law function
.powerlaw <- function(x, C, alpha) {
  C * x^(-alpha)
}
```


```{r}
library(HiCcompare)

data('HMEC.chr22')
data('NHEK.chr22')
hic.table <- create.hic.table(HMEC.chr22, NHEK.chr22, chr = 'chr22')
# Plug hic.table into hic_loess()
result <- hic_loess(hic.table, Plot = TRUE)
# perform difference detection
diff.result <- hic_diff(result, diff.thresh = 'auto', Plot = TRUE)

hic.table <- diff.result

######
library(poweRlaw)
# calculate A
hic.table[, A := (adj.IF1 + adj.IF2) / 2]
fit <- conlnorm$new(hic.table$A)
# fit$setXmin(1)
est <- estimate_xmin(fit)
fit$setXmin(est)
est <- estimate_pars(fit)
fit$setPars(est)


fit <- conlnorm$new(hic.table$A)
est <- estimate_xmin(fit)
fit$setXmin(est)
est <- estimate_pars(fit)
fit$setPars(est)
plot(fit)
lines(fit, col=2)

fit2 <- conpl$new(hic.table$A)
est <- estimate_xmin(hic.table$A)
fit2$setXmin(est)
est <- estimate_pars(hic.table$A)
fit2$setPars(est)
plot(fit2)
lines(fit2, col=2)

# test if powerlaw fit is good using KS
# bs_p <- bootstrap_p(fit, no_of_sims = 10, threads = 1)


# library(fitdistrplus)
library(igraph)
fit_power_law(hic.table$A)
length(hic.table$A[hic.table$A < 4000])
fit_power_law(hic.table$A[hic.table$A < 4000])

# fit power law by deleting most extreme values
fit_powerlaw = function(x) {
  x = na.omit(x) # remove any NAs
  fit = power.law.fit(x) # get initial fit
  pval = fit$KS.p
  # loop until fit is good
  while(pval <= 0.05) {
    x = x[-which(x == max(x))]
    fit = power.law.fit(x)
    pval = fit$KS.p
  }
  return(fit)
}


fit <- fit_powerlaw(hic.table$A)

 .powerlaw(hic.table$A, fit$xmin, fit$alpha)


.powerlaw(7, fit$xmin, fit$alpha)
# between 6 and 7 the powerlaw fit hits 1. values > 7 are below 1 and and x increases to inf the value returned goes to 0

# test multiplying powerlaw result by p-value
adj.p <- hic.table$p.value * (1 + .powerlaw(hic.table$A, fit$xmin, fit$alpha))
hist(adj.p)

plaw <- .powerlaw(hic.table$A, fit$xmin, fit$alpha)
hist(plaw)

cbind(hic.table$A, hic.table$p.value, adj.p)[1:100,]
cbind(hic.table$A, hic.table$p.value, adj.p)[2000:2100,]

MD.plot2(hic.table$adj.M, hic.table$D, hic.table$p)
MD.plot2(hic.table$adj.M, hic.table$D, adj.p)
# that isn't going to work. Makes pretty much everything super significant

# could we weight fold changes based on average expression before performing permutation test instead?
# find midpoint of log A, this corresponds to weight of 1. A < midpoint downweight M and A > midpoint upweight M??


# maybe find change point on log A centered bell curve then use that as the multiplier. Things on tails either thrown 
# out for left tail or multiplied by 1 for right tail


# try weighting by raw difference
hic.table <- diff.result
weight <- (abs(hic.table$adj.IF1 - hic.table$adj.IF2) / max(c(hic.table$adj.IF1, hic.table$adj.IF2)))

hic.table[, weight := weight]
hic.table[, adj.p := p.value * weight]
hic.table
```

# Test function on different datasets

## set up prostate data
```{r}
library(GenomicRanges)
core_path <- "D:/3D_DNA/rickman/rickman_lib1/"
man_path <- "D:/3D_DNA/manuscript/prostate_analysis/"

chrs <- paste0('chr', 1:22)
chrs <- c(chrs, 'chrX')

# read in data for 1 mb
ERG_lib1 <- list()
GFP_lib1 <- list()
for(i in 1:22) {
  ERG_lib1[[i]] <- read.table(paste0(core_path, 'ERG_lib1_chr', i, '.1mb.txt'), header = FALSE)
  GFP_lib1[[i]] <- read.table(paste0(core_path, 'GFP_lib1_chr', i, '.1mb.txt'), header = FALSE)
}
ERG_lib1[[23]] <- read.table(paste0(core_path, 'ERG_lib1_chr', 'X', '.1mb.txt'), header = FALSE)
GFP_lib1[[23]] <- read.table(paste0(core_path, 'GFP_lib1_chr', 'X', '.1mb.txt'), header = FALSE)

cnv <- read.table('D:/3D_DNA/rickman/cnv_out.bed', header = TRUE)

# cnv_rows <- which(cnv$ERG_lib1_1_2.hicup == -2 | cnv$ERG_lib1_1_2.hicup == 2 | cnv$GFP_lib1_1_2.hicup == -2 | cnv$GFP_lib1_1_2.hicup == 2)
# exclude <- cnv[cnv_rows, 2:4]
# cnv <- cnv[cnv_rows,]
cnv.gi <- GRanges(seqnames = cnv$chromosome, ranges = IRanges(start = cnv$start, end = cnv$end))
colnames(cnv)[1] <- 'chr'
exclude <- rbind(cnv, hg19_blacklist)
#exclude <- hg19_blacklist
exclude.gi <- GRanges(seqnames = exclude$chr, ranges = IRanges(start = exclude$start, end = exclude$end))

lib1_tables <- mapply(create.hic.table, ERG_lib1, GFP_lib1, chrs, SIMPLIFY = FALSE, MoreArgs = list(scale = FALSE, include.zeros = FALSE,
                                                                                                    exclude.regions = exclude, 
                                                                                                    exclude.overlap = 0.2))


lib1_tables <- total_sum(lib1_tables)


lib1_tables <- hic_loess(lib1_tables, Plot = TRUE, span = NA)
lib1_tables <- hic_diff(lib1_tables, Plot = TRUE, diff.thresh = NA, iterations = 1000)


```

## test p-value adjustment

### chr22 test data
```{r}
data('HMEC.chr22')
data('NHEK.chr22')
hic.table <- create.hic.table(HMEC.chr22, NHEK.chr22, chr = 'chr22')
# Plug hic.table into hic_loess()
result <- hic_loess(hic.table, Plot = TRUE)
# perform difference detection
diff.result <- hic_diff(result, diff.thresh = 'auto', Plot = TRUE)

hic.table <- diff.result

adjust_pval(hic.table)
```

### prostate data
```{r}
chr1 <- lib1_tables[[1]]
chr1 <- adjust_pval(chr1)

par(mfrow = c(1,2))
MD.plot2(chr1$M, chr1$D, chr1$p.value)
MD.plot2(chr1$M, chr1$D, chr1$adj.p)

# Powerlaw Xmin = 7.29600338922691 Powerlaw alpha = 84.4788481904521
```


# Test M weighting instead

```{r}
hist(log2(hic.table$A))
hist(log2(chr1$A))

which(chr1$A == median(chr1$A))
which(log2(chr1$A) == median(log2(chr1$A)))

fit_power_law(chr1$A[chr1$A < quantile(chr1$A, 0.75)])

quantile(chr1$A)
hist(chr1$A[chr1$A > 1 & chr1$A < 6.14])
plot(density(chr1$A[chr1$A > 1 & chr1$A < 16]))

# scale density to be on (0,1)
tmpA <- (chr1$A[chr1$A > 1 & chr1$A < 16]) / max(chr1$A[chr1$A > 1 & chr1$A < 16])*5
plot(density(tmpA))

# plot(density(log2(chr1$A[chr1$A > 1 & chr1$A < 16])))

density_A <- density(chr1$A[chr1$A > 1 & chr1$A < 16])
A_pdf <- approxfun(density_A)
A_pdf(3)

density_A$x[which.max(density_A$y)]

1 - A_pdf(1.7)
1 - A_pdf(10)
1 - (A_pdf(2) / max(density_A$y))
q <- seq(1.7, 16, length.out = 1000)
plot(1 - A_pdf(q) ~ q)
plot(1 - (A_pdf(q) / max(density_A$y)) ~ q)
```

So maybe for values above 75% quantile on A, we trust so no down weighting to these values of M
For A values < 1 we dont trust at all so should not be significant, weight their M to 0

## test function

```{r}
chr1 <- weight_M(chr1, quant = 0.75)
MD.plot2(chr1$adj.M, chr1$D)

chr1 <- hic_diff(chr1, diff.thresh = 'auto', Plot = TRUE)


hic.table <- weight_M(hic.table, quant = 0.1)
MD.plot2(hic.table$adj.M, hic.table$D)

new_tables <- lapply(lib1_tables, weight_M)

pdf(file = "D:/3D_DNA/manuscript/prostate_analysis/M_weighting_prostate.pdf")
new_tables <- hic_diff(new_tables, Plot = TRUE, diff.thresh = 'auto', iterations = 1000)
dev.off()
```


### what happens if we pool all chrs together for M weighting

```{r}
library(data.table)
lib1_tables <- mapply(create.hic.table, ERG_lib1, GFP_lib1, chrs, SIMPLIFY = FALSE, MoreArgs = list(scale = FALSE, include.zeros = FALSE,
                                                                                                    exclude.regions = exclude, 
                                                                                                    exclude.overlap = 0.2))


lib1_tables <- total_sum(lib1_tables)


lib1_tables <- hic_loess(lib1_tables, Plot = TRUE, span = NA)
pool_tables <- rbindlist(lib1_tables)
pool_tables <- weight_M(pool_tables)
pool_tables <- hic_diff(pool_tables, Plot = TRUE, diff.thresh = 'auto')

pool_tables[p.value < 0.05 & D > 200,]
```


# test on brain 50kb data

```{r}
amyg_bed <- read.table('D:/brain/data/amyg/50000/amyg_50000_abs.bed')
amyg_mat <- read.table('D:/brain/data/amyg/50000/amyg_50000.matrix')
dplfc_bed <- read.table('D:/brain/data/dplfc/50000/dplfc_50000_abs.bed')
dplfc_mat <- read.table('D:/brain/data/dplfc/50000/dplfc_50000.matrix')
amyg_data <- hicpro2bedpe(amyg_mat, amyg_bed)
dplfc_data <- hicpro2bedpe(dplfc_mat, dplfc_bed)

amyg_dplfc <- mapply(create.hic.table, amyg_data$cis[-23], dplfc_data$cis[-23], MoreArgs = list(exclude.regions = NA, exclude.overlap = 0.2,
                                                                                      scale = FALSE), SIMPLIFY = FALSE)

amyg_dplfc <- total_sum(amyg_dplfc)

# amyg_dplfc <- hic_loess(amyg_dplfc, Plot = TRUE)
# amyg_dplfc <- hic_diff(amyg_dplfc, Plot = TRUE, diff.thresh = 'auto')

bchr1 <- amyg_dplfc[[1]]
bchr22 <- amyg_dplfc[[22]]

bchr22 <- hic_loess(bchr22, Plot = TRUE)
bchr22 <- weight_M(bchr22)

hist(bchr22$A)
hist(bchr22$A[bchr22$A > 1 & bchr22$A < quantile(bchr22$A, 0.75)])
hic_diff(bchr22, Plot = TRUE)

bchr1 <- hic_loess(bchr1, Plot = TRUE)
bchr1 <- weight_M(bchr1)
bchr1 <- hic_diff(bchr1, Plot = TRUE)
```

