---
title: "`HiCdiff` Vignette"
author: "John Stansfield, Mikhail Dozmorov"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HiCdiff Usage Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```
# Introduction

`HiCdiff` provides functions for the joint normalization and detection of differential chromatin interactions between two or multiple Hi-C datasets. `HiCdiff` operates on processed Hi-C data in the form of chromosome-specific chromatin interaction matrices. It accepts three-column tab-separated text files storing chromatin interaction matrices in a sparse matrix format (see [Creating the hic.table object](#hic_table)). Functions to convert popular Hi-C data formats (`.hic`, `.cool`) to sparse format are available (see ?cooler2sparse). `HiCdiff` differs from other packages that attempt to compare Hi-C data in that it works on processed data in chromatin interaction matrix format instead of pre-processed sequencing data. In addition, `HiCdiff` provides a non-parametric method for the joint normalization and removal of biases between two Hi-C datasets for the purpose of comparative analysis. `HiCdiff` also provides a simple yet robust permutation method for detecting differences between Hi-C datasets. 

The `hic_loess` function outputs normalized chromatin interactions for both matrices ready for the comparative analysis. The `hic_diff` function performs the comparative analysis and outputs genomic coordinates of pairs of regions detected as differentially interacting, interaction frequencies, the difference and the corresponding permutation p-value.

# How to use `HiCdiff`

## Installation

Install `HiCdiff` from Bioconductor.

```{r, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("HiCdiff")
library(HiCdiff)
```


## Getting Hi-C Data

You will need processed Hi-C data in the form of sparse upper triangular matrices or BEDPE files in order to use `HiCdiff`. Data is available from several sources and two examples for downloading and extracting data are listed below. If you have full Hi-C contact matrices you can convert them to sparse upper triangular format using the full `full2sparse` function as shown in [additional functions](#addfunc)

### Extracting data from `.hic` files

Hi-C data is available from several sources and in many formats. `HiCdiff` is built to work with the sparse upper triangular matrix format popularized by the lab of Erez Lieberman-Aiden [http://aidenlab.org/data.html](http://aidenlab.org/data.html).  If you already have Hi-C data either in the form of a sparse upper triangular matrix or a full contact matrix you can skip to the [Creating the hic.table object](#hic_table) section. If you obtain data from the Aiden Lab in the `.hic` format you will need to first extract the matrices that you wish to compare. 

  1. Download the `straw` software from [https://github.com/theaidenlab/straw/wiki](https://github.com/theaidenlab/straw/wiki) and install it.
  2. Use `straw` to extract a Hi-C sparse upper triangular matrix. An example is below:
  
Say we downloaded the `GSE63525_K562_combined_30.hic ` file from GEO [https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525)  

To extract the raw matrix corresponding to chromosome 22 at 500kb resolution we would use the following command within the terminal

  `./straw NONE GSE63525_K562_combined_30.hic  22 22 BP 500000 > K562.chr22.500kb.txt`

This will extract the matrix from the `.hic` file and save it to the `K562.chr22.500kb.txt` text file, in the sparse upper triangular matrix format. See more examples on on how to use `straw` at [https://github.com/theaidenlab/straw/wiki/CPP#running](https://github.com/theaidenlab/straw/wiki/CPP#running). Straw requires several inputs for the extraction of data from a `.hic` file.

`<NONE/VC/VC_SQRT/KR> <hicFile(s)> <chr1>[:x1:x2] <chr2>[:y1:y2] <BP/FRAG> <binsize>`

The first argument is the normalization method. For use in `HiCdiff` you want the raw data so you should selected `NONE`. The second argument is the `.hic` file name. Next is the chromosome numbers of the matrix you want. For an intrachromosomal contact map both should be the same as in the above example. If you want a matrix of interchromosomal interactions you can use different chromosomes i.e. interactions between chromosome 1 and chromosome 2 (Note that `HiCdiff` is only meant to be used on intrachromosomal interactions at this point in development). The next argument is whether you want basepair or fragment files. For `HiCdiff` use `BP`. The final argument is the binsize of the matrix (the resolution). To extract a matrix at a resolution of 1MB enter `10000000`. Typical bin sizes are 1MB, 500KB, 100KB, 50KB, 5KB, 1KB. Note that most matrices with resolutions higher than 100KB (i.e. matrices with resolutions of 1KB - 50KB) are typically too sparse (due to insufficient sequencing coverage) for analysis in `HiCdiff`. 

From here we can just import the matrix into R as you would normally for any tab-delimited file.

  3. Import the data into R `K562.chr22 <- read.table('K562.chr22.500kb.txt', header=FALSE)`
  4. Repeat these steps for any other Hi-C dataset that you wish to compare to the first dataset using  `HiCdiff`, then proceed to [Creating the `hic.table` object](#hic_table).
 
### Extracting data from `.cool` files

The `cooler` software, [http://cooler.readthedocs.io/en/latest/index.html](http://cooler.readthedocs.io/en/latest/index.html), allows access to a large collection of Hi-C data. The cooler index [ftp://cooler.csail.mit.edu/coolers](ftp://cooler.csail.mit.edu/coolers) contains Hi-C data for `hg19` and `mm9` from many different sources. To use data in the `.cool` format in `HiCdiff` follow these steps:

  1. Download and install `cooler` from [http://cooler.readthedocs.io/en/latest/index.html](http://cooler.readthedocs.io/en/latest/index.html)
  2. Download a `.cool` file from the cooler index [ftp://cooler.csail.mit.edu/coolers](ftp://cooler.csail.mit.edu/coolers).
  3. Say we downloaded the `Dixon2012-H1hESC-HindIII-allreps-filtered.1000kb.cool` file. See `cooler dump --help` for data extraction options. To extract the contact matrix we use the following commands in the terminal:  
  `cooler dump --join Dixon2012-H1hESC-HindIII-allreps-filtered.1000kb.cool > dixon.hESC.1000kb.txt`
  4. Read in the text file as you would any tab-delimited file in R  
  `hesc1000kb <- read.table("dixon.hESC.1000kb.txt", header = FALSE)`
  5. Convert to a sparse upper triangular matrix using the `HiCdiff::cooler2sparse` function.  
  `sparse <- cooler2sparse(hesc1000kb) `
  6. Repeat the steps for another Hi-C dataset that you wish to compare to the first dataset then proceed to [Creating the `hic.table` object](#hic_table).
  
## Creating the `hic.table` object {#hic_table}

### Sparse upper triangular format

A sparse matrix format represents a relatively compact and human-readable way to store pair-wise interactions. It is a tab-delimited text format containing three columns: "region1" - a start coordinate (in bp) of the first region, "region2" a start coordinate of the second region, and "IF" - the interaction frequency between them (IFs). Zero IFs are dropped (hence, the _sparse_ format). Since the full matrix of chromatin interactions is symmetric, only the upper triangular portion, including the diagonal, is stored.

If you have two full Hi-C contact matrices you can convert them to sparse upper triangular matrices using the `HiCdiff::full2sparse` function. Once you have two sparse matrices you are ready to create a `hic.table` object. This will be illustrated using the included sparse matrices at 500kb resolution for chromosome 22 from the HMEC and NHEK cell lines.

```{r, warning=FALSE, message=FALSE}
library(`HiCdiff`)
# load the data
data("HMEC.chr22")
data("NHEK.chr22")
head(HMEC.chr22)
```

Now that we have 2 sparse upper triangular matrices we can create the `hic.table` object. `create.hic.table` requires the input of 2 sparse matrices and the chromosome name.

```{r}
# create the `hic.table` object
chr22.table <- create.hic.table(HMEC.chr22, NHEK.chr22, chr = 'chr22')
head(chr22.table)
```

We can also list multiple `hic.tables` in order to utilize parallel computing. First we create another `hic.table` as was done above. Then we combine these two `hic.tables` into a list.

```{r}
# create list of `hic.table` objects
data("HMEC.chr10")
data("NHEK.chr10")

# create the `hic.table` object
chr10.table <- create.hic.table(HMEC.chr10, NHEK.chr10, chr = 'chr10')
hic.list <- list(chr10.table, chr22.table)
head(hic.list)
```

The `hic.table` object contains a summary of the differences between the two matrices. "IF1" and "IF2" correspond to interaction frequencies in the first and second matrices, "D" is the unit distance (length of each unit is equivalent to the resolution of the data, e.g., 500kb), "M" is the $log_2(IF2)-log_2(IF1)$ difference.

### BEDPE format

A `hic.table` object can also be created using data in the 7 column BEDPE format. An example of BEDPE data for the HMEC dataset used above is shown below.

```{r, echo=FALSE}
HMEC.chr22_BEDPE <- chr22.table[, 1:7, with=FALSE]
NHEK.chr22_BEDPE <- chr22.table[, c(1:6, 8), with=FALSE]
head(HMEC.chr22_BEDPE)
```

To create a `hic.table` object using BEDPE data is very similar to using data in the sparse upper triangular format.

```{r}
bed.hic.tab <- create.hic.table(HMEC.chr22_BEDPE, NHEK.chr22_BEDPE)
head(bed.hic.tab)
```

### InteractionSet/GRanges format

A `hic.table` can also be created using an `InteractionSet` object. Simply enter two `InteractionSets` representing two Hi-C matrices into the `create.hic.table` function and they will be converted to the proper format. See the `InteractionSet` vignette for creating `InteractionSet` objects [here](https://bioconductor.org/packages/release/bioc/vignettes/InteractionSet/inst/doc/interactions.html)

```{r}
data("hmec.IS")
data("nhek.IS")
head(hmec.IS)
IS.hic.tab <- create.hic.table(hmec.IS, nhek.IS)
```

To shorten computational time, or if one is only interested in a subsection of a Hi-C matrix, one may wish to use a subset of the `hic.table` object. Use the `subset.dist` or `subset.index` options, see help for the `create.hic.table` function.

## Joint Normalization

Now that you have created a `hic.table` object you can jointly normalize your two Hi-C matrices. The `hic_loess` function has many options and can accept a single `hic.table` or a list of hic.tables. If for example you wish to perform joint normalization for every chromosome on two cell lines while utilizing parallel computing you can create a list containing the hic.tables for each chromosome. 

To change the degree of the polynomial for the loess joint normalization you can utilize the `degree` option, default is 1 (linear regression). A user-defined span, or the amount of data used to build the loess model, can also be set with the `span` option. However if `span = NA` (the default) the automatic smoothing parameter selection process will run and determine the optimal span for the data. The type of selection process can be changed using the `loess.criterion` option. Available settings are `gcv` (the default) for generalized cross-validation or `aicc` for Akaike information criterion. The loess automatic smoothing parameter selection uses a customized version of the `fANCOVA::loess.as` function. For more information on parameter selection please see the `fANCOVA` reference manual. It is recommended to use the default settings.

`hic_loess` can utilize the `BiocParallel` package to perform parallel computing and lower computation time. The `parallel` option (FALSE by default) will only speed up computation if a list of hic.tables is entered into the function, i.e., it parallelizes processing of several chromosome-specific matrices. This is useful for performing joint normalization for every chromosome between two Hi-C datasets. For more information on `BiocParallel` see the reference manual [here.](http://bioconductor.org/packages/release/bioc/vignettes/BiocParallel/inst/doc/Introduction_To_BiocParallel.pdf)

The basis of `HiCdiff` rests on the novel concept termed the MD plot. The MD plot is similar to the MA plot or the Bland-Altman plot. $M$ is the log2 difference between the interaction frequencies from the two datasets. $D$ is the unit distance between the two interacting regions. Loess is performed on the data after it is represented in the MD coordinate system. To visualize the MD plot the `Plot` option can be set to TRUE. 

If you wish to detect differences between the two Hi-C datasets immediately following joint normalization you can set the `check.differences` option to TRUE. However, if you only want to jointly normalize the data for now keep this option set to FALSE. The difference detection process will be covered in the next section.  

```{r, fig.width=7, fig.height=4}
# Jointly normalize data for a single chromosome
hic.table <- hic_loess(chr22.table, Plot = TRUE)
head(hic.table)
```

```{r, message=FALSE}
# Multiple hic.tables can be processed in parallel by entering a list of hic.tables
hic.list <- hic_loess(hic.list, parallel = TRUE)
```

The `hic_loess` joint normalization function extends the `hic.table` with the adjusted interaction frequencies, adjusted "M", and the "mc" correction factor.

## Difference Detection

Difference detection also makes use of the MD plot. The jointly normalized datasets are once again plotted on the MD plot. Difference detection takes place on a per-unit-distance basis using a permutation test. Permutations are broken into blocks for each unit distance. See methods section of _INSERT PAPER HERE_ for more details. Difference detection can be performed at the same time as joint normalization if the `check.differences` option in `hic_loess` is set to TRUE. Otherwise a `hic.table` object output from the `hic_loess` function can be input into the `hic_diff` function for difference detection. 

`hic_diff` accepts a `hic.table` object or a list of hic.table objects output from `hic_loess`. `hic_diff` can also utilize the `BiocParallel` package the same way as the `hic_loess` loess function. The same limits and uses apply as stated above. If the `Plot` option is set to TRUE, an MD plot will be displayed showing the normalized MD plot along with coloring based on the significance of each difference. The dotted lines on this MD plot represent the `diff.thresh` threshold. This threshold is automatically calculated as 2 standard deviations of M from 0 on the MD plot by default but can be set by the user or turned off. This option helps to deal with false positives by restricting the regions that can be labelled as significant when they fall inside the threshold. The threshold can be thought of as a range in which differences do not exceed the level of the expected noise in the Hi-C data. The function will return a hic.table or a list of hic.tables with an additional column containing the p-value for the significance of the difference between the IFs in each cell of the two Hi-C matrices. 

If you wish to visualize the p-values in the form of a Hi-C contact map you can use the `sparse2full` function to transform the results to a contact matrix with the p-values inserted instead of IFs as illustrated in the example below.

```{r, fig.width=7, fig.height=6, warning=FALSE}
# input hic.table object into hic_diff
hic.table <- hic_diff(hic.table, Plot = TRUE)

# input a list of hic.tables
hic.list <- hic_diff(hic.list)

# transform results into contact map of p-values
library(pheatmap)
m <- sparse2full(hic.table, hic.table = TRUE, column.name = 'p.value') # convert to full matrix
fc <- sparse2full(hic.table, hic.table = TRUE, column.name = 'M')
pheatmap(-log10(m) * sign(fc), cluster_rows = FALSE, cluster_cols = FALSE)
```


## Converting `HiCdiff` results to `InteractionSet` objects

If after running `hic_loess` or `hic_diff` on your data you wish to perform additional analyses which require the `GRanges` object class you can convert the results using the `make_InteractionSet` function. This function will produce an `InteractionSet` object with the genomic ranges contained in the `hic.table` along with several metadata files containing the additional information produced by `hic_loess` or `hic_diff`.

```{r}
IntSet <- make_InteractionSet(hic.table)
```



## Simulating Hi-C Data

`HiCdiff` includes functions for simulating Hi-C data. The `hic_simulate` function allows you to simulate two Hi-C matrices with added bias and true differences at a specified fold change. As an example we will simulate two matrices with 250 true differences added at a fold change of 4.

```{r fig.width=7, fig.height=4}
number_of_unitdistances <- 100 # The dimensions of the square matrix to be simualted
number_of_changes       <- 250 # How many cells in the matrix will have changes
i.range <- sample(1:number_of_unitdistances, number_of_changes, replace = TRUE) # Indexes of cells to have controlled changes
j.range <- sample(1:number_of_unitdistances, number_of_changes, replace = TRUE) # Indexes of cells to have controlled changes

sim_results <- hic_simulate(nrow = number_of_unitdistances, medianIF = 50000, sdIF = 14000, powerlaw.alpha = 1.8, fold.change = 4, i.range = i.range, j.range = j.range, Plot = TRUE, alpha = 0.1)
```

The results of the simulation are saved in a list.

```{r}
names(sim_results)
```

TPR is the true positive rate, SPC is the specificity, pvals is a vector of the p-values for every cell in the matrices, hic.table is the resulting hic.table object after `hic_loess` and `hic_diff` have been applied to the simulated data, true.diff is a table for the cells that had the specified fold change applied to them, truth is a vector of 0's and 1's indicating if a cell had a true difference applied - this is useful for input into ROC packages, sim.table is the simulated data in a hic.table object before being scaled, normalized, and analyzed for differences.

## Additional Functions {#addfunc}

`HiCdiff` contains some additional functions that may be useful. 

If you do not choose to show the MD plots when you initially run `hic_loess` or `hic_diff` you can use the `MD.plot1` or `MD.plot2` functions. `MD.plot1` will create a side by side MD plot showing before and after loess normalization. Enter your original M and D vectors along with the M correction factor, `mc`, calculated by `hic_loess`. 

```{r, fig.height=4, fig.width=7}
MD.plot1(M = hic.table$M, D = hic.table$D, mc = hic.table$mc)
```

`MD.plot2` will create a standard MD plot with optional colouring based on p-value. Just enter an M and D vector and a p-value vector if desired.

```{r, fig.height=4, fig.width=7}
# no p-value coloring
MD.plot2(M = hic.table$adj.M, D = hic.table$D)

# p-value coloring
MD.plot2(M = hic.table$adj.M, D = hic.table$D, hic.table$p.value)
```

There are two matrix transformation functions included. `sparse2full` will transform a sparse upper triangular matrix to a full Hi-C matrix. `full2sparse` will transform a full Hi-C matrix to sparse upper triangular format.

```{r}
full.NHEK <- sparse2full(NHEK.chr22)
full.NHEK[1:5, 1:5]

sparse.NHEK <- full2sparse(full.NHEK)
head(sparse.NHEK)
```



`KRnorm` will perform Knight-Ruiz normalization on a Hi-C matrix. Just enter the full Hi-C matrix to be normalized.

```{r}
KR.NHEK <- KRnorm(full.NHEK)
```


`SCN` will perform Sequential Component Normalization on a Hi-C matrix. Just enter the full Hi-C matrix to be normalized.

```{r}
SCN.NHEK <- SCN(full.NHEK)
```


`MA_norm` will perform MA normalization on a `hic.table` object.

```{r, message=FALSE}
result <- MA_norm(hic.table, Plot = TRUE)
```

