---
title: Estimation of the power-law dependence between the $log_{10}-log_{10}$ interaction
  frequencies and distance between interacting regions
author: "John Stansfield, Mikhail Dozmorov"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
bibliography: ../3D_DNA-references.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 5.5, fig.height = 3) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r libraries}
library(HiCcompare)
library(igraph)
library(ggplot2)
library(gridExtra)
```

# Introduction

To estimate the parameters for simulating individual Hi-C matrices, Hi-C data from Gm12878 cell line [@Rao:2014aa] were used (Supplementary Table 1, [GSE63525](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525)). The first dataset was obtained with the DpnII restriction enzyme, while the second dataset was obtained with the MboI enzyme. Data from chromosome 1 was used at resolutions of 1Mb, 500kb, 100kb, and 50kb. The data were converted in a sparse upper triangular matrix format (see `HiCcompare-vignette.Rmd` for details). Additional Gm12878 Hi-C data from chrs 1, 18, and 19 at 1Mb resolution, cut using the DpnII and MboI enzymes were also included. 

```{r}
# load data
githubURL <- "https://github.com/dozmorovlab/HiCdiff/raw/supplemental/Supplemental_data/S2_File_data.RData"
load(url(githubURL))



# Pairwise matrices as `hic.table` objects
table.1mb   = create.hic.table(S2.dpnii.chr1, S2.mbol.chr1, 'chr1', scale = F, include.zeros = T)
table.500kb = create.hic.table(S2.dpnii.500kb, S2.p.500kb,  'chr1', scale = F, include.zeros = T)
table.100kb = create.hic.table(S2.dpnii.100kb, S2.p.100kb,  'chr1', scale = F, include.zeros = T)
table.50kb  = create.hic.table(S2.dpnii.50kb, S2.p.50kb,    'chr1', scale = F, include.zeros = T)


chr1.tab  = create.hic.table(S2.dpnii.chr1,  S2.mbol.chr1,  chr = 'chr1')
chr18.tab = create.hic.table(S2.dpnii.chr18, S2.mbol.chr18, chr = 'chr18')
chr19.tab = create.hic.table(S2.dpnii.chr19, S2.mbol.chr19, chr = 'chr19')
```

# Median IF vs. distance dependence

First, we estimate the power-law approximation of the dependence between interaction frequencies (IFs) and distance [@Lieberman-Aiden:2009aa; @Sanborn:2015aa; @Ay:2015aa; @Fudenberg:2016aa; @Nagano:2015aa] by fitting the power-law estimate and assessing the fit using Kolmogorov-Smirnov test. Because IFs at larger distances do not fit the power-law well (Supplementary Figure 1) the outlier values are iteratively removed, starting from the largest distances, until the Kolmogorov-Smirnov test results indicate the power-law fit is adequate. The $\alpha$ power-law parameter can then be used to approximate the decay of median IF with distance.

The power-law approximation can be affected by multiple factors, e.g., the resolution of the data, the enzymes used to obtain the data, the chromosomal differences (e.g., chromosome length (chromosome 1 being the longest), gene density (gene-poor chromosome 18 and gene-dense chromosome 19)). 

## Effect of Resolution

Here the fit to the power-law of median IF at each distance is tested for chromosome 1 from GM12878 cell line using cutting enzymes DpnII and MboI at 1MB, 500KB, 100KB, 50KB resolution, respectively.

The table shows the output of `fitdistplus::power.law.fit` function. Key variables to note are `alpha` - the power of the $C*x^{-alpha}$ power-law formula, and `KS.p` - p-value of the Kolmogorov-Smirnov test, larger p-value means that the power-law fit is adequate. The plots represents the log10(median IF) and log10(Distance), one plot for each cutting enzyme.

```{r}
kb1000 = aggregate(table.1mb$IF1,   by = list(table.1mb$D),   FUN = median)
kb500  = aggregate(table.500kb$IF1, by = list(table.500kb$D), FUN = median)
kb100  = aggregate(table.100kb$IF1, by = list(table.100kb$D), FUN = median)
kb50   = aggregate(table.50kb$IF1,  by = list(table.50kb$D),  FUN = median)
```

```{r}
mkb1000 = aggregate(table.1mb$IF2,   by = list(table.1mb$D),   FUN = median)
mkb500  = aggregate(table.500kb$IF2, by = list(table.500kb$D), FUN = median)
mkb100  = aggregate(table.100kb$IF2, by = list(table.100kb$D), FUN = median)
mkb50   = aggregate(table.50kb$IF2,  by = list(table.50kb$D),  FUN = median)
```

```{r}
chr1  = aggregate(chr1.tab$IF1,  by = list(chr1.tab$D),  FUN = median)
chr18  = aggregate(chr18.tab$IF1,  by = list(chr18.tab$D),  FUN = median)
chr19 = aggregate(chr19.tab$IF1, by = list(chr19.tab$D), FUN = median)
# chr15 = aggregate(chr15.tab$IF1, by = list(chr15.tab$D), FUN = median)
# chr22 = aggregate(chr22.tab$IF1, by = list(chr22.tab$D), FUN = median)

mchr1  = aggregate(chr1.tab$IF2,  by = list(chr1.tab$D),  FUN = median)
mchr18  = aggregate(chr18.tab$IF2,  by = list(chr18.tab$D),  FUN = median)
mchr19 = aggregate(chr19.tab$IF2, by = list(chr19.tab$D), FUN = median)
# mchr15 = aggregate(chr15.tab$IF2, by = list(chr15.tab$D), FUN = median)
# mchr22 = aggregate(chr22.tab$IF2, by = list(chr22.tab$D), FUN = median)
```


```{r}
# function to iteratively remove the observations at the largest distance until the power law is met
fit_powerlaw = function(tab) {
  tab = na.omit(tab) # remove any NAs
  fit = power.law.fit(tab$x) # get initial fit
  pval = fit$KS.p
  # loop until fit is good
  while(pval <= 0.05) {
    tab = tab[-nrow(tab),]
    fit = power.law.fit(tab$x)
    pval = fit$KS.p
  }
  return(tab)
}
```

```{r}
kb1000 = fit_powerlaw(kb1000)
kb500  = fit_powerlaw(kb500)
kb100  = fit_powerlaw(kb100)
kb50   = fit_powerlaw(kb50) 

mkb1000 = fit_powerlaw(mkb1000) 
mkb500  = fit_powerlaw(mkb500)  
mkb100  = fit_powerlaw(mkb100)  
mkb50   = fit_powerlaw(mkb50)  

chr1  = fit_powerlaw(chr1)
chr18  = fit_powerlaw(chr18)
chr19 = fit_powerlaw(chr19)


mchr1  = fit_powerlaw(mchr1)
mchr18  = fit_powerlaw(mchr18)
mchr19 = fit_powerlaw(mchr19)
```

```{r}
# fucntion for making ggplots 

sdplot = function(D, IF, title = 'Median IF vs D plot', linfit = FALSE) {
  dat = data.frame(D = D, IF = IF)
  p = ggplot(dat, aes(x = D, y = IF)) + geom_point() + ggtitle(title) 
  if (linfit) {
    p = p + geom_smooth(method='lm', se = FALSE) # add the linear fit for the powerlaw
  } else {
    p = p + scale_x_continuous(limits = c(0, 50))
  }
  return(p)
}
```

### 1 MB

```{r}
t1 = power.law.fit(kb1000$x) 
t2 = power.law.fit(mkb1000$x) 
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1 = sdplot(log10(kb1000$Group.1),  log10(kb1000$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mkb1000$Group.1), log10(mkb1000$x), title = 'MboI log10(IF) vs log10(D)',  linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### 500KB
```{r}
t1 = power.law.fit(kb500$x)   
t2 = power.law.fit(mkb500$x) 
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1 = sdplot(log10(kb500$Group.1),  log10(kb500$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mkb500$Group.1), log10(mkb500$x), title = 'MboI log10(IF) vs log10(D)',  linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### 100KB
```{r}
t1 = power.law.fit(kb100$x)  
t2 = power.law.fit(mkb100$x) 
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1 = sdplot(log10(kb100$Group.1),  log10(kb100$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mkb100$Group.1), log10(mkb100$x), title = 'MboI log10(IF) vs log10(D)',  linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### 50KB
```{r}
t1 = power.law.fit(kb50$x) 
t2 = power.law.fit(mkb50$x) 
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1 = sdplot(log10(kb50$Group.1),  log10(kb50$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mkb50$Group.1), log10(mkb50$x), title = 'MboI log10(IF) vs log10(D)',  linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### Summary

The power-law seems to fit the data fairly well over the varying resolutions once the outliers at the furthest distances are removed. The estimates of $\alpha$ range from 1.73 to 2.07. 


## Effect of Chromosome

Here the fit to the power-law of median IF at each distance is tested for chromosome 1, 18, 19 from GM12878 cell line using cutting enzymes DpnII and MboI at 1MB resolution. 

As above, the table shows the output of `fitdistplus::power.law.fit` function. Key variables to note are `alpha` - the power of the $C*x^{-alpha}$ power-law formula, and `KS.p` - p-value of the Kolmogorov-Smirnov test, larger p-value means that the power-law fit is adequate. The plots represents the log10(median IF) and log10(Distance), one plot for each cutting enzyme.

### Chr 1

```{r}
t1 = power.law.fit(chr1$x)  
t2 = power.law.fit(mchr1$x)
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1= sdplot(log10(chr1$Group.1),  log10(chr1$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mchr1$Group.1), log10(mchr1$x), title = 'MboI log10(IF) vs log10(D)', linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### Chr 18

```{r}
t1 = power.law.fit(chr18$x)  
t2 = power.law.fit(mchr18$x) 
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1 = sdplot(log10(chr18$Group.1),  log10(chr18$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mchr18$Group.1), log10(mchr18$x), title = 'MboI log10(IF) vs log10(D)',  linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### Chr 19

```{r}
t1 = power.law.fit(chr19$x)  
t2 = power.law.fit(mchr19$x) 
t3 = rbind(t1, t2)
t3 = cbind(Enzyme = c('DpnII', 'MboI'), t3)
rownames(t3) = NULL
t3  %>% as.data.frame() %>% pander()
```

```{r}
p1 = sdplot(log10(chr19$Group.1),  log10(chr19$x),  title = 'DpnII log10(IF) vs log10(D)', linfit = T)
p2 = sdplot(log10(mchr19$Group.1), log10(mchr19$x), title = 'MboI log10(IF) vs log10(D)',  linfit = T)
grid.arrange(p1, p2, ncol=2)
```

### Summary

The power-law fits the data over varying chromosomes fairly well after the outliers are removed. $\alpha$ ranges from 1.7 to 2.05. For the simulation methods it seems that choosing an $\alpha$ between 1.7 and 2.05 should be adequate. 

# References
